---
title: Tables & Fonts
author: Adam Filipowicz
date: '2020-05-06'
slug: tables-fonts
categories: []
tags:
  - Data Visualization
summary: 'A brief look at creating tables and using fonts in R (note: tables do not display well using the dark theme).'
authors: []
external_link: ''
image:
  placement: 1
  caption: ''
  focal_point: 'Smart'
  preview_only: no
url_code: ''
url_pdf: ''
url_slides: ''
url_video: ''
slides: ''
output: 
  html_document:
    df_print: kable
---

# Lab 5 - Tables & Fonts

```{r echo=FALSE}
library(pnwflights14)
library(tidyverse)
library(kableExtra)
```


## Tables

Note that the [pnwflights14 package](https://github.com/ismayc/pnwflights14) includes several other data sets that may be of use (mappings from two-letter airline codes to airline names, weather conditions, etc.).

```{r echo=FALSE}
data("flights", package = "pnwflights14")
data("weather", package = "pnwflights14")
data("airports", package = "pnwflights14")
data("airlines", package = "pnwflights14")
```

### Departure track records

Which airlines had the best and worst track records of on-time departures in each month? Is it different between PDX and SEA?

Let's start by selecting just the month, departure delay, carrier, and origin columns in the flights dataset. We'll also append a name column on from the airlines dataset, using the shared carrier column.

```{r}
departure_times <- flights %>% 
  select(month, dep_delay, carrier, origin) %>% 
  left_join(airlines, by = "carrier")
glimpse(departure_times)
```

Alright, looks good. Now, we'll group by month, name and origin and get the mean delay time.

```{r}
mean_departure_times <- departure_times %>% 
  group_by(month, name, origin) %>% 
  summarise(mean_delay=mean(dep_delay, na.rm = TRUE))
glimpse(mean_departure_times)
```

Alright, let's setup kable for a table.

```{r}
options(knitr.table.format = "html")
options(scipen = 1, digits = 2)
```

And actually make the table.

```{r}
worst_departure_times <- mean_departure_times %>%
  group_by(month, origin) %>% top_n(1, mean_delay)
best_departure_times <- mean_departure_times %>% 
  group_by(month, origin) %>% top_n(-1, mean_delay)
bw_departure_times <- bind_rows(best_departure_times, worst_departure_times) %>% 
  arrange(month, mean_delay)
departure_times_cols <- c("Month", "Airline", "Origin", "Delay (minutes)")
bw_departure_times %>% 
  kable(col.names = departure_times_cols) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

My approach for tables is "simpler is better". I gave the table some stripes to break up the monotony since it is a rather long table, I set it so that it wouldn't take up the whole page, and then forced it to the left side. Things are arranged chronologically first, and then in each month we get the best and worst delay times for PDX and SEA, with the 2nd column telling us which airline it was. Simple and quite readable I think.

### Service from PDX

What cities have the most service from Portland (defined as number of flights from Portland)? Which have the least?

We'll need the flights dataset again, but this time select out just origin and destination columns. We'll filter the origin so we're just looking at PDX flights, then get a count based on destinations. Then, to make the final table more readable, we'll rename the "dest" column to "faa", join by that column to get the airport names in, select just the name of the airports and the counts ("n"), and then rename things for our table:

```{r}
cities_serviced <- flights %>%
  select(origin, dest) %>% 
  filter(origin == "PDX") %>% 
  count(dest) %>% 
  rename(faa = dest) %>% 
  left_join(airports, by = "faa") %>% 
  select(name, n) %>% 
  rename(Airport = name, Flights = n)
glimpse(cities_serviced)
```

Alright, that looks ready, let's make the table:

```{r}
top_cities_serviced <- cities_serviced %>%
  arrange(desc(Flights)) %>% 
  head(n=5)
bottom_cities_serviced <- cities_serviced %>% 
  arrange(desc(Flights)) %>% 
  tail(n=5)
tb_cities_serviced <- bind_rows(top_cities_serviced, bottom_cities_serviced)
tb_cities_serviced %>% 
  kable() %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

I've chosen here to display just the top 5 and bottom 5 airports. Styling is similar to the table above. This gives the viewer a nice quick way to see which cities receive the most or least amount of flights in all of 2014.

### Which aircraft log the most distance?

Finding busy aircraft, with business defined as having the most distance travelled. We'll pull out "tailnum" and "distance" columns from flights, group by the tail number, and then sum up the distances for each. Then, we'll arrange the summed distances in a descending fashion, and get just the top 10. Renaming the columns of the table, we'll go ahead and make it using similar styling as the tables above (adding in a title this time.)

```{r}
aircraft <- flights %>%
  select(tailnum, distance) %>% 
  group_by(tailnum) %>% 
  summarise(Distance = sum(distance))
busy_aircraft <- aircraft%>%
  arrange(desc(Distance)) %>% 
  head(n=10)
busy_aircraft_cols <- c("Aircraft", "Distance (miles)")
busy_aircraft %>% 
  kable(col.names = busy_aircraft_cols, caption = "Busiest (most distance travelled) aircraft in 2014") %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

Cool! I think that gives a really quick, non-cluttered way to see which aircraft are the busiest.

## Fonts

```{r echo=FALSE}
library(extrafont)
```

In the original assignment I had trouble gettin extrafonts working nicely with RStudio Cloud. Let's get a list of the fonts available to us using a local setup, though.

```{r}
fonttable() %>% select(FamilyName) %>% distinct()
```

Many more than what I had using the cloud version. Let's just see what we can do with those.

We're going to use the demo plot below to show off three different fonts.

```{r}
demo_plt <- ggplot(iris, aes(Sepal.Length, Sepal.Width, color=Species)) + 
  geom_point(size=3) +
  scale_color_discrete(labels=c("Setosa", "Versicolor", "Virginica")) + 
  labs(x="Sepal Length (cm)", y="Sepal Width (cm)", title="Fisher's Irises") 
demo_plt
```

### Serif

```{r}
demo_plt + theme(text = element_text(family="Palatino Linotype"))
```

I don't really like this one for graph labels, but I do think it would work pretty well for the body of text, especially when printed out.

### Sanserif

```{r}
demo_plt + theme(text = element_text(family="Lato"))
```

For a graph like this, this is my favorite font of the three. It shows up well on a screen and is very simple to read. It works well for everything from the title to the figure legend.

### Display

```{r}
demo_plt + theme(text = element_text(family="Broadway"))
```

This one is actually really fun looking to me. Maybe a little too stylized for something like figure labels, but I can see using it for headers or anywhere you want to give the text a little pop.
